---
title: "Smoking and Drinking"
author: "Brian High"
date: "2/19/2016"
output: 
    ioslides_presentation:
        fig_width: 8.5
        fig_height: 3.5
        keep_md: yes
        smaller: yes
---

## SQL Examples: Smoking and Drinking

This is a demo of some basic SQL `SELECT` queries using BRFSS data from: 
http://www.cdc.gov/brfss/. 

We have downloaded the data for each respondent for the years 2005 through 2014.

This dataset has 4,379,516 rows and 999 columns.

This dataset will be too large to fit in RAM memory for most desktop and laptop 
computers.

When exported as a TSV file, this file is 13 GB. When loaded into R as a data 
table, the memory consumed is almost 33 GB.

Instead, we have imported the data into a MySQL database table. We have indexed
the table on survey year and state to improve performance.

The CDC has provided a 
[codebook](http://www.cdc.gov/brfss/annual_data/2014/pdf/codebook14_llcp.pdf) 
for use in understanding variables and codes.

In particular, we will focus on tobacco use and alcohol consumption in 
the state of Washington.

```{r set_options, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(tidy=FALSE, cache=TRUE)
```

## Connect to MySQL Database

We will connect to the `localhost` and `brfss` database using an `anonymous` 
account.

```{r message=FALSE, cache=FALSE}
library(RMySQL)

con <- dbConnect(MySQL(), 
                 host="localhost", 
                 username="anonymous", 
                 password="Ank7greph-", 
                 dbname="brfss")
```

It's generally a *bad* idea to put your connection credentials in your script,
and an even *worse* idea to publish these on Github. *Don't be like me!*

```{r}
if (file.exists("con.R")) source("con.R")
```

A lesser evil is to put them in a separate file that you keep secure and private.

Even better would be to configure your system to prompt you for the password.

## Table Size

You can find the number of rows and columns, as well as table and index size,
with a few simple queries.

```{r}
sql <- "SELECT COUNT(*) AS rows FROM brfss;"
rs <- dbGetQuery(con, sql)
cat(rs$rows, "rows")
sql <- "SELECT * FROM brfss LIMIT 1;"
rs <- dbGetQuery(con, sql)
cat(ncol(rs), "columns")
sql <- "SHOW TABLE STATUS IN brfss;"
rs <- dbGetQuery(con, sql)
cat(sum(rs[1, c("Data_length", "Index_length")]) / 1024 / 1024 / 1024, "GB")
```

## Count Respondents per Year

Usually, aggregating functions like `COUNT()` are used with `GROUP BY`. Let's
count the number of respondents per year in Washington state (`53`). `ORDER BY` 
is for sorting.

```{r}
sql <- "SELECT IYEAR AS Year, COUNT(*) AS Respondents 
        FROM brfss 
        WHERE X_STATE = 53 
        GROUP BY IYEAR 
        ORDER BY IYEAR;"
rs <- dbGetQuery(con, sql)
rs
```

## Respondents per Education Level

Since 2015 has very few respondents in our dataset, we will use as 2014 as
the most recent year. To get a idea of our sample, we look at education level.

```{r}
sql <- "SELECT X_EDUCAG AS Education, COUNT(*) AS Respondents 
        FROM brfss 
        WHERE IYEAR = 2014 AND X_STATE = 53 
        GROUP BY X_EDUCAG 
        ORDER BY X_EDUCAG;"
rs <- dbGetQuery(con, sql)
rs
```

The education level (`X_EDUCAG`) is an integer from 1-4 (or 9 meaning 
"Don't know", "Missing", etc.). What trend do we see here?

## Count Smokers by Education Level

We can check the `USENOW3` variable to see if the survey respondent is
a smoker or not. A value of `1` (Every day) or `2` (Some days) means 
"is a smoker".

```{r}
sql <- "SELECT X_EDUCAG AS Education, 
        COUNT(USENOW3) AS Smokers 
        FROM brfss 
        WHERE IYEAR = 2014 
              AND X_STATE = 53 
              AND X_EDUCAG <= 4 
              AND (USENOW3 = 1 OR USENOW3 = 2) 
        GROUP BY X_EDUCAG 
        ORDER BY X_EDUCAG;"
rs <- dbGetQuery(con, sql)
rs
```

Since the number of respondents varies by education level, we will want to 
calculate "prevalence" using respondent totals by education level.

## Count Smokers by Education Level

We can get a count of smokers and total respondents per education level in one 
query by using the `IF()` function within the `COUNT()` function.

```{r}
sql <- "SELECT X_EDUCAG AS Education, 
        COUNT(*) AS Respondents, 
        COUNT(IF(USENOW3 = 1 OR USENOW3 = 2, 1, NULL)) AS Smokers 
        FROM brfss 
        WHERE IYEAR = 2014 
              AND X_STATE = 53 
              AND X_EDUCAG <= 4 
        GROUP BY X_EDUCAG 
        ORDER BY X_EDUCAG;"
rs <- dbGetQuery(con, sql)
rs
```

The `IF()` condition `USENOW3 = 1 OR USENOW3 = 2` was taken from the `WHERE` 
clause. We had to make this change so that `COUNT(*)` counts all respondents.

## Smoking Prevalence by Education Level

We use functions from the `dplyr` package to calculate smoking prevalence. This
is the number of smokers as a fraction of respondents for each education level.

```{r message=FALSE}
library(dplyr)
rs %>% group_by(Education) %>% 
    mutate(Smoking.Prevalence=Smokers/Respondents) -> smokers
smokers
```

## Relabel Education Level

Now, we relabel the codes for education level to meaningful text strings. We 
abbreviate the "Value Label" text descriptions from the codebook as follows.

```{r}
edu.labels <- c("some school", "high school grad", 
                "some college", "college grad")
smokers$Education <- factor(smokers$Education, levels=1:4, labels=edu.labels)
smokers
```

## Smoking Prevalence by Education Level

```{r message=FALSE}
library(ggplot2)
ggplot(data=smokers, aes(x=Education, y=Smoking.Prevalence, fill=Education)) +
    geom_bar(stat="identity")
```

```{r echo=F, results='hide'}
dev.off()
```

## Count Smokers by Education and Year

How has smoking changed from 2011 to 2014?

```{r}
sql <- "SELECT IYEAR AS Year, X_EDUCAG AS Education, 
        COUNT(*) AS Respondents, 
        COUNT(IF(USENOW3 = 1 OR USENOW3 = 2, 1, NULL)) AS Smokers
        FROM brfss 
        WHERE (IYEAR >= 2011 AND IYEAR <= 2014)
              AND X_STATE = 53 
              AND X_EDUCAG <= 4 
        GROUP BY IYEAR, X_EDUCAG 
        ORDER BY IYEAR, X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
rs %>% group_by(Year, Education) %>% 
    mutate(Smoking.Prevalence=Smokers/Respondents) -> smokers
```

## Smoking by Education and Year

```{r message=FALSE}
smokers$Education <- factor(smokers$Education, levels=1:4, labels=edu.labels)
smokers$Year <- factor(smokers$Year)
ggplot(data=smokers, aes(x=Education, y=Smoking.Prevalence, fill=Year)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black")
```

```{r echo=F, results='hide'}
dev.off()
```

## Count Drinkers by Education Level

The `DRNKANY5` variable stores a value indicating if the survey respondent has 
consumed an alcoholic drink in the past 30 days. We will use this value to 
indicate if the survey respondent is currently a drinker or not. A value of
`1` means "is a drinker". Again, we will just look at Washington state in 2014.

```{r}
sql <- "SELECT X_EDUCAG AS Education, 
        COUNT(*) AS Respondents, 
        COUNT(IF(DRNKANY5 = 1, 1, NULL)) AS Drinkers 
        FROM brfss 
        WHERE IYEAR = 2014
              AND X_STATE = 53 
              AND X_EDUCAG <= 4 
        GROUP BY X_EDUCAG 
        ORDER BY X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
```

## Drinking Prevalence by Education Level

Again, using `dplyr`, we can calculate drinking prevalence.

```{r}
rs %>% group_by(Education) %>% 
    mutate(Drinking.Prevalence=Drinkers/Respondents) -> drinkers
drinkers$Education <- factor(drinkers$Education, levels=1:4, labels=edu.labels)
drinkers
```

## Drinking Prevalence by Education Level

```{r}
ggplot(data=drinkers, aes(x=Education, y=Drinking.Prevalence, fill=Education)) +
    geom_bar(stat="identity")
```

```{r echo=F, results='hide'}
dev.off()
```

## Count Drinkers by Education and Year

Let's see how drinking compares from 2011 to 2014.

```{r}
sql <- "SELECT IYEAR AS Year, X_EDUCAG AS Education, 
        COUNT(*) AS Respondents, 
        COUNT(IF(DRNKANY5 = 1, 1, NULL)) AS Drinkers 
        FROM brfss 
        WHERE (IYEAR >= 2011 AND IYEAR <= 2014)
              AND X_STATE = 53 
              AND X_EDUCAG <= 4 
        GROUP BY IYEAR, X_EDUCAG 
        ORDER BY IYEAR, X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
rs %>% group_by(Year, Education) %>% 
    mutate(Drinking.Prevalence=Drinkers/Respondents) -> drinkers
```

## Drinking by Education and Year

```{r}
drinkers$Education <- factor(drinkers$Education, levels=1:4, labels=edu.labels)
drinkers$Year <- factor(drinkers$Year)
ggplot(data=drinkers, aes(x=Education, y=Drinking.Prevalence, fill=Year)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black")
```

```{r echo=F, results='hide'}
dev.off()
```

## Why so many queries?

We could have retrieved all of the data for the previous plots with a single
query:

```{r}
sql <- "SELECT IYEAR AS Year, X_EDUCAG AS Education, 
        COUNT(*) AS Respondents, 
        COUNT(IF(USENOW3 = 1 OR USENOW3 = 2, 1, NULL)) AS Smokers, 
        COUNT(IF(DRNKANY5 = 1, 1, NULL)) AS Drinkers 
        FROM brfss 
        WHERE (IYEAR >= 2011 AND IYEAR <= 2014)
              AND X_STATE = 53 
              AND X_EDUCAG <= 4 
        GROUP BY IYEAR, X_EDUCAG 
        ORDER BY IYEAR, X_EDUCAG;"

rs <- dbGetQuery(con, sql)
rs %>% group_by(Year, Education) %>% 
    mutate(Smoking=Smokers/Respondents, 
           Drinking=Drinkers/Respondents) -> consumers
consumers$Education <- factor(consumers$Education, levels=1:4, 
                              labels=edu.labels)
consumers$Year <- factor(consumers$Year)
```

From this dataframe, just subset as needed to produce tables and plots.

## Smoking and Drinking Prevalence

```{r results='asis'}
consumers %>% head(8) %>% kable(digits=3)
```

## Smoking and Drinking in Long Format

To facilitate plotting, we will want to group by consumption type. To do this,
we will need to convert the data structure from "wide" to "long" format. The
`gather()` function of the `tidyr` package makes this easy.

```{r}
library(tidyr)
consumers <- consumers %>% 
    select(Year, Education, Smoking, Drinking) %>% 
    gather(key=Factor, value=Prevalence, -Year, -Education)
head(consumers, 8)
```

## Smoking and Drinking Prevalence

```{r}
ggplot(data=consumers, aes(x=Year, y=Prevalence, group=Factor, color=Factor)) + 
    geom_line() + facet_grid(Factor ~ Education, scales="free_y")
```

```{r echo=F, results='hide'}
dev.off()
```

## Compare other Variables

Now that you know how to query the database, compare other variables, such as:

- Smoking and drinking by income (`X_INCOMG`) or race (`X_RACE`)
- Binge drinking (`X_RFBING5`) by gender (`SEX`) or age (`X_AGE80`)
- BMI category (`X_BMI5CAT`) and exercise (`EXERANY2`) or sleep (`SLEPTIM1`)
- Internet use (`INTERNET`) and mental health (`MENTHLTH`) or 

## Close Database Connection

Once we are done with the database, we can close the connection to it.

```{r}
# Close connection
dbDisconnect(con)
```