---
title: "Smokers and Drinkers"
author: "Brian High"
date: "2/18/2016"
output: 
    ioslides_presentation:
        fig_width: 8.5
        fig_height: 3.5
        keep_md: yes
        smaller: yes
        cache: true
---

## SQL Examples: Smokers and Drinkers

This is a demo of some basic SQL `SELECT` queries using BRFSS data 
from: http://www.cdc.gov/brfss/. 

Only one table is used, so we will not need and "JOIN" statements. 

The CDC has provided a 
[codebook](http://www.cdc.gov/brfss/annual_data/2013/pdf/codebook13_llcp.pdf) 
for use in understanding variables and codes.

In particular, we will focus on tobacco use and alcohol consumption in 
the state of Washington.

## Connect to MySQL Database

We will connect to the `localhost` and `brfss` database using an `anonymous` 
account.

```{r}
library(RMySQL)

con <- dbConnect(MySQL(), 
                 host="localhost", 
                 username="anonymous", 
                 password="Ank7greph-", 
                 dbname="brfss")
```

## Count Smokers by Education Level

We can check the `USENOW3` variable to see if the survey respondent is
a smoker or not. A value of `1` (Every day) or `2` (Some days) means 
"is a smoker". We will restrict the year (`IYEAR`) to `2014` and the 
state (`X_STATE`) to `53` ("Washington"). The education level 
(`X_EDUCAG`) is an integer from 1-4 (or 9 meaning "Don't know", "Missing", etc.).

```{r}
sql <- "SELECT X_EDUCAG AS Education, count(USENOW3) AS Smokers 
        FROM brfss 
        WHERE IYEAR = 2014 
              AND X_STATE = 53 
              AND (USENOW3 = 1 OR USENOW3 = 2) 
              AND X_EDUCAG <= 4 
        GROUP BY X_EDUCAG 
        ORDER BY X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
rs
```

## Relabel Education Level

We will relabel the codes for education level to have meaningful labels. We will
abbreviate the "Value Label" text descriptions from the codebook as follows.

```{r}
edu.labels <- c("some school", "high school grad", 
                "some college", "college grad")
rs$Education <- factor(rs$Education, levels=1:4, labels=edu.labels)
rs
```

## Smokers by Education Level

```{r}
library(ggplot2)
ggplot(data=rs, aes(x=Education, y=Smokers, fill=Education)) +
    geom_bar(stat="identity")
```

```{r echo=F, results='hide'}
dev.off()
```

## Smokers by Education Level and Year

How has smoking changed from 2012 to 2014?

```{r}
sql <- "SELECT IYEAR AS Year, X_EDUCAG AS Education, count(USENOW3) AS Smokers
        FROM brfss 
        WHERE (IYEAR >= 2012 AND IYEAR <= 2014)
              AND X_STATE = 53 
              AND (USENOW3 = 1 OR USENOW3 = 2) 
              AND X_EDUCAG <= 4 
        GROUP BY IYEAR, X_EDUCAG 
        ORDER BY IYEAR, X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
rs$Education <- factor(rs$Education, levels=1:4, labels=edu.labels)
```

## Smokers by Education Level and Year

```{r}
rs$Year <- factor(rs$Year)
ggplot(data=rs, aes(x=Education, y=Smokers, fill=Year)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black")
```

```{r echo=F, results='hide'}
dev.off()
```

## Count Drinkers by Education Level

The `DRNKANY5` variable stores a value indicating if the survey respondent has 
consumed an alcoholic drink in the past 30 days. We will use this value to 
indicate if the survey respondent is currently a drinker or not. A value of
`1` means "is a drinker". Again, we will just look at Washington state in 2013.

```{r}
sql <- "SELECT X_EDUCAG AS Education, count(DRNKANY5) AS Drinkers 
        FROM brfss 
        WHERE IYEAR = 2014
              AND X_STATE = 53 
              AND DRNKANY5 = 1 
              AND X_EDUCAG <= 4 
        GROUP BY X_EDUCAG 
        ORDER BY X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
rs$Education <- factor(rs$Education, levels=1:4, labels=edu.labels)
rs
```

## Drinkers by Education Level

```{r}
ggplot(data=rs, aes(x=Education, y=Drinkers, fill=Education)) +
    geom_bar(stat="identity")
```

```{r echo=F, results='hide'}
dev.off()
```

## Drinkers by Education Level and Year

Let's see how drinking compares from 2012 to 2014.

```{r}
sql <- "SELECT IYEAR AS Year, X_EDUCAG AS Education, count(DRNKANY5) AS Drinkers 
        FROM brfss 
        WHERE (IYEAR >= 2012 AND IYEAR <= 2014)
              AND X_STATE = 53 
              AND DRNKANY5 = 1 
              AND X_EDUCAG <= 4 
        GROUP BY IYEAR, X_EDUCAG 
        ORDER BY IYEAR, X_EDUCAG DESC;"

rs <- dbGetQuery(con, sql)
rs$Education <- factor(rs$Education, levels=1:4, labels=edu.labels)
```

## Drinkers by Education Level and Year

```{r}
rs$Year <- factor(rs$Year)
ggplot(data=rs, aes(x=Education, y=Drinkers, fill=Year)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black")
```

```{r echo=F, results='hide'}
dev.off()
```

## Smokers and Drinkers by Year

We can compare the number of smokers and drinkers by year in one query using
the SQL `IF()` function within our `COUNT()` function.

```{r}
sql <- "SELECT IYEAR as Year,
        COUNT(IF(USENOW3 = 1 OR USENOW3 = 2, 1, NULL)) AS Smokers,
        COUNT(IF(DRNKANY5 = 1, 1, NULL)) AS Drinkers
        FROM brfss
        WHERE IYEAR <= 2014 AND X_STATE = 53
        GROUP BY IYEAR 
        ORDER BY IYEAR;"

rs <- dbGetQuery(con, sql)
```

This saves us from otherwise having to perform two separate queries and then 
having to merge the results. SQL does all of this for us.

## Smokers and Drinkers by Year

Now we can show the smokers and drinkers side-by-side in "wide" format.

```{r}
rs$Year <- factor(rs$Year)
consumers <- rs
consumers
```

## Smokers and Drinkers in Long Format

To facilitate plotting, we will want to group by consumption type. To do this,
we will need to convert the data structure from "wide" to "long" format. The
`gather()` function of the `tidyr` package makes this easy.

```{r}
library(tidyr)
consumers <- gather(consumers, key=Type, value=Count, -Year)
head(consumers)
```

## Smokers and Drinkers by Year

```{r}
ggplot(data=consumers, aes(x=Year, y=Count, group=Type, color=Type)) +
    geom_line()
```

```{r echo=F, results='hide'}
dev.off()
```

## Close Database Connection

Since we are done with the database, we can close the connect to it.

```{r}
# Close connection
dbDisconnect(con)
```